Global spirals in self-gravitating dead zones in protoplanetary discs

### 2014-10-31 paper/model.tex
major change in paper - new idea - ang mom exchange with background because of temp gradient

The non-axisymmetric surface density can be visualised with
\begin{align}
  \Delta \Sigma = \frac{\Sigma - \Sigma_{0}}{\Sigma_{0}}, 
\end{align}
while an individual mode as
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
	\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$.

The global mode amplitude is 
\begin{align}
  A_m = \frac{|C_m|}{M_{d0}}
\end{align}
where $M_{d0}$ is the initial disc mass and 
\begin{align}
  C_m = 2\pi\int_{R_1}^{R_2} \Sigma_m RdR, 
\end{align}
with integration limits chosen later.

The time evolution of the $m^\mathrm{th}$ mode can be characterized by
assuming 
$\Sigma_m\propto\exp{(\ii \sigma t)}$, where
$\sigma = \omega - \ii\gamma$ is the complex frequency, $\omega$ is
the real frequency and $\gamma$ is the growth rate.
We estimate the growth rate as
\begin{align}
  \gamma = \frac{d\ln{A_m}}{dt},
\end{align}
and we estimate the real frequency by calculating the pattern speed
$\Omega_p = \omega/m$,
\begin{align}
  \Omega_p = \frac{d\phi_\mathrm{max}}{dt},
\end{align}
where $\phi_\mathrm{max}$ is the azimuthal co-ordinate of
$\mathrm{max}(\Delta\Sigma_m)$. 
%\begin{align}
%  \mathrm{max}\left\{2 \real\left[ \Sigma_m \exp{(\ii m
%        \phi)}\right]\right\},

%other methods possible too

\subsection{Angular momentum decomposition}
The total disc angular momentum in $R\in[R_1,R_2]$ is
\begin{align}
  J &= \int_{R_1}^{R_2}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
  &= 2\pi\int_{R_1}^{R_2}R\Sigma_0 v_{\phi0} R dR \notag\\ 
  &\phantom{=}+
  \sum_{m=1}^\infty2\pi\int_{R_1}^{R_2} 2R\real\left[\Sigma_m v_{\phi
      m}^*\right] RdR, \notag\\
  &= \sum_{m=0}^\infty \int_{R_1}^{R_2}  j_m(R) dR = \sum_{m=0}^\infty J_m. 
\end{align}
where $^*$ denotes complex conjugation, and we idenfity $J_m$ as the angular momentum 
associated with the $m^\mathrm{th}$ mode, and $j_m$ the associated
angular momentum per unit length.

\subsection{Linear density waves}\label{wkb}
We will also use linear density wave theory to interpret
simulation data. Here, we follow conventions in \cite{shu91} and quote
results therein. The dispersion relation for tightly-wound density
waves of the form 
$\exp{[\ii(\omega t - m \phi + kR)]}$ in a razor-thin
isothermal disc is 
\begin{align}\label{dispersion}
  (\omega - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $\kappa^2 = R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency, 
and $k$ is a real wavenumber such that $|kR|\gg1$.

@@

% We use Eq. \ref{wavenumber} to set trailing long waves as
% perturbations, i.e. $k/k_T = -2Q^{-2}\left[1 - \sqrt{1 -
%     Q^2(1-\nu^2)}\right]$.  
% We only apply the perturbations where $Q^{-2}(1-\nu^2)<1$.

% The amplitude of the surface density and velocity perturbations are
% related by
% \begin{align}
%   &\left(\omega - m\Omega\right)\delta\Sigma + k\Sigma\delta v_R= 0,\\
%   &\frac{\kappa^2}{2\Omega} \delta v_R + \ii (\omega - m\Omega)\delta
%   v_\phi=0. 
% \end{align}
% We set $\delta\Sigma = 10^{-3}\Sigma$ and $(\delta v_R, \delta
% v_\phi)$ accordingly.

% %Having chosen the wave frequency and set the wavenumber according to
% %the above,

>>>>

The normalised surface density associated with $m$ is
\begin{align}
  \Delta\Sigma_m = \frac{2}{\Sigma_{00}} \real\left[\Sigma_m \exp{(\ii
      m\phi)}\right]
\end{align}
where $\Sigma_{00} = \Sigma_0(t=0)$. The time evolution of the
$m^\mathrm{th}$ mode can be characterized by 
assuming $\Sigma_m\propto\exp{(-\ii \sigma t)}$, where
$\sigma = \omega + \ii\gamma$ is the complex frequency, $\omega$ is
the real frequency and $\gamma$ is the growth rate. The total
non-axisymmetric surface density is 
\begin{align}
  \Delta\Sigma = \frac{\Sigma - \Sigma_0}{\Sigma_0}. 
\end{align}

\subsubsection{Angular momentum decomposition}
The total disc angular momentum in $R\in[R_1,R_2]$ is
\begin{align}
  J &= \int_{R_1}^{R_2}\int_0^{2\pi} \Sigma Rv_\phi RdRd\phi \notag\\
  &= 2\pi\int_{R_1}^{R_2}R\Sigma_0 v_{\phi0} R dR \notag\\ 
  &\phantom{=}+
  \sum_{m=1}^\infty2\pi\int_{R_1}^{R_2} 2R\real\left[\Sigma_m v_{\phi
      m}^*\right] RdR, \notag\\
  &= \sum_{m=0}^\infty \int_{R_1}^{R_2}  j_m(R) dR = \sum_{m=0}^\infty J_m. 
\end{align}
where $^*$ denotes complex conjugation, and we idenfity $J_m$ as the 
angular momentum associated with disc structure with
$m^\mathrm{th}$-fold symmetry, and $j_m$ the associated angular
momentum per unit length. We will use $J_m$ to monitor angular momentum 
conservation for the disc as a whole.

\subsection{Linear density waves}\label{wkb}
In addition to the above diagnostics for disc evolution, we
will also use linear density wave theory to interpret simulation
results.

\subsubsection{Global angular momentum conservation for linear
  perturbations} 
It can be shown that linear perturbations with $\phi$-dependence in the form
$\exp{\ii m\phi }$ satisfies angular momentum conservation
in the form 
\begin{align} 
  \frac{\p \jlin}{\p t} + \nabla\cdot\bm{F} = \tbg, 
\end{align}
where
\begin{align}
  \jlin \equiv
  -\frac{m\Sigma}{2}\imag\left(\bm{\xi}^*\cdot\frac{\p\bm{\xi}}{\p
      t} + \Omega \hat{\bm{k}}\cdot\bm{\xi}\times\bm{\xi}^* + \ii
    m \Omega |\bm{\xi}|^2\right) 
\end{align}
is the angular momentum density, $\bm{\xi}$ is the Lagragian
displacement, and $\bm{F}$ is the angular momentum flux
consisting of a Reynolds stress and a gravitational stress. Explicit
expressions for $\bm{\xi}$ and $\bm{F}$ can be found in e.g.,
\cite{papaloizou85} and \cite{lin11} respectively. In the above
expressions, $\Sigma$ and $\Omega$ correspond to the unperturbed disc.

The term $\tbg$ is 
\begin{align}
  \tbg \equiv
  -\frac{m}{2}\imag\left(\Sigma_m\xi_R^\prime\frac{dc_s^2}{dR}\right), 
\end{align}
and arises because we have adopted a locally isothermal equation of
state. In a barotropic fluid, such as a strictly isothermal disc,
$\tbg$ vanishes and the angular momentum associated with
the perturbation is conserved, provided boundary fluxes vanish.

However, as noted in \cite{lin11}, if there is an imposed
temperature gradient, as is the disc models considered in this paper,
then $\tbg\neq 0$ in general, which corresponds to a torque
exerted by the background disc on the perturbation.

We remark that the perturbed angular momentum density $\jlin$ has been
defined to satisfy a conservation law. This differs from the 
empirical angular momentum decomposition above.

%each m satisfies its own equation

\subsubsection{Local results}
\cite{shu91} gives the dispersion relation for tightly-wound density
waves of the form $\exp{[\ii(\omega t - m \phi + kR)]}$ in a razor-thin
isothermal disc as  
\begin{align}\label{dispersion}
  (\omega - m\Omega)^2 = \kappa^2 + k^2c_s^2 - 2\pi G \Sigma |k|, 
\end{align}
where $\kappa^2 = R^{-3}\p_R(R^4\Omega^2)$ is the square of the epicycle frequency, 
and $k$ is a real wavenumber such that $|kR|\gg1$.

@@

### 2014-10-31 paper/results.tex
major change in paper - new idea - ang mom exchange with background because of temp gradient

\begin{figure}
  % scale=0.41
  \includegraphics[scale=.41]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the fiducial
    FARGO simulation. The perturbation
    relative to $t=0$ in 2D is shown in units of the
    initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom}} 
\end{figure}

\subsection{Non-linear initial perturbation}
In order to focus on the one-arm spiral that emerges in the above
simulation, we here analyse a run with only an initial $m=1$
perturbation in the dead zone. Specifically, we set
\begin{align}
  &\Sigma \to \Sigma_0(R) + \delta\Sigma_0(R) + 2\Sigma_1(R)\cos{\phi},\\
  &v_\phi  \to v_{\phi 0}(R) + 2v_{\phi 1}(R)\cos{\phi},
\end{align}
for $R\in[R_{d1},R_{d2}]$, where
\begin{align}
  &\delta\Sigma_0 =
  -\delta^2\times\Sigma_0(R_{d1})\frac{R_{d1}}{R}\sin^3{\left[\frac{2\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},\\ 
  &\Sigma_1 =
  \delta\times\Sigma_0\sin{\left[\frac{\pi\left(R-R_{d1}\right)}{R_{d2}-R_{d1}}\right]},  
\end{align}
with $\delta = 10^{-3}$, and
\begin{align}
  v_{\phi 1} = -\frac{v_{\phi 0}\delta\Sigma_0}{2\Sigma_1}. 
\end{align}
This perturbs the dead zone structure while conserving total mass and
angular momentum. The $m=1$ perturbation has an associated negative
angular momentum (as observed in the previous simulation) at the
expense of increasing the angular momentum associated with the
background disc. %local exchange for simplicity 
%actually doesn't matter if we set plus or minus
Note that the perturbation to the axisymmetric surface density is
$O(\delta^2)$, whereas that in $m=1$ is $O(\delta)$. Since $Q>1$
everywhere in the disc, there is no risk of axisymmetric gravitational
instability.

While the above form of initial perturbations is artificial, we find 
the spiral that forms have similar properties (see below) to that found in the
previous simulation. However, the above perturbations have the
advantage of producing cleaner results without complications from
higher $m$ modes, and the required simulation time for the one-arm
spiral to emerge is significantly reduced.

Fig. shows the evolution of angular momenta.

>>>>

\begin{figure}
  % scale=0.41
  \includegraphics[width=\linewidth]{figures/nonaxi_evol_ang_fargo}
  \caption{Evolution of angular momentum components in the fiducial
    FARGO simulation. The perturbation
    relative to $t=0$ in 2D is shown in units of the
    initial total angular momentum $J_\mathrm{ref}$.\label{2d_angmom}} 
\end{figure}
